```{r setup, include=FALSE}
# packages to be installed 
install_and_load_packages <- function() {
  # List of packages to be installed and loaded
  packages <- c(
    "labelled",
    "rstatix",
    "ggpubr",
    "GGally",
    "car",
    "Epi",
    "lme4",
    "lmerTest",
    "emmeans",
    "multcomp",
    "geepack",
    "ggeffects",
    "gt",
    "readxl",
    "data.table",
    "tidyverse",
    "summarytools",
    "ggplot2",
    "dplyr",
    "lubridate",
    "readr",
    "stringr",
    "tibble",
    "naniar",
    "NHANES",
    "forcats",
    "moments"
  )
  
  # Install packages that are not already installed
  installed_packages <- installed.packages()
  for (pkg in packages) {
    if (!(pkg %in% installed_packages[, "Package"])) {
      install.packages(pkg)
    }
  }
  
  # Load all the packages
  lapply(packages, library, character.only = TRUE)
}

# Run the function to install and load all packages
install_and_load_packages()

#Relevant libraries for use in this data 
library(labelled)   # This is used to labeling data
library(rstatix)    # This is used for summary statistics
library(ggpubr)     # This is used for convenient summary statistics and plots
library(GGally)     # This is used for advanced plot
library(car)        # This is used for useful for anova/wald test
library(Epi)        # This is used for easy getting CI for model coef/pred
library(lme4)       # This is used for linear mixed-effects models
library(lmerTest)   # This is used for test for linear mixed-effects models
library(emmeans)    # This is used for marginal means
library(multcomp)   # This is used for CI for linear combinations of model coef
library(geepack)    # This is used for generalized estimating equations
library(ggeffects)  # This is used for marginal effects, adjusted predictions
library(gt)         # This is used for nice tables
library(readxl)     # This is used for reading Excel files
library(data.table) # This is used for fast data manipulation
library(tidyverse)  # for data manipulation and visualization
library(summarytools) # This is used for detailed summary statistics
library(ggplot2)    # This is used for data visualization
library(dplyr)      #This is used  for data manipulation
library(lubridate)  # This is used for date-time manipulation
library(readr)      # This is used for reading data
library(stringr)    # This is used for string manipulation
library(tibble)     # This is used for modern data frames
library(naniar)     # This is used for handling missing data
library(NHANES)     # This is used for accessing NHANES data
library(forcats)    # This is used for working with categorical variables
library(moments)
```

# Step 1
# This is the importation of the "longi_new_data" big dataframe in the last analysis into this work because that is the last step before the data is converted Into the longitudinal format. Here we have to consider all the partitions within the dataframe we will name them now and in the course of this modeling we might consider new partitions directly or by grouping when nessesary. 
# First the three major dataframe is imported which are difference in days, binary_df which are subset of the main dataframe longi_new_data
# We have already identified the Response variable, therefore we can subset the Response from the main dataframe which is the longi_new_data to arrive at the X variables which are binary_df, difference_in_days , the interventions and the patient data, so if we remove the Response , binary_df, difference_in_days from the longi_new_data we arrive at a dataframe that is the interventions and the patient demograpics , in understanding how this model work we might at some point work our way back to the interventions and the patients demographic to see how the changes we make her will impact the model. 
```{r}
longi_new_data<-read.csv("D:/mimiproject/longi_new_data.csv", header = TRUE, stringsAsFactors = FALSE)
difference_in_days<-read.csv("D:/mimiproject/difference_in_days.csv", header = TRUE, stringsAsFactors = FALSE)
binary_df<-read.csv("D:/mimiproject/binary_df.csv", header = TRUE, stringsAsFactors = FALSE)
# 
```

```{r}
longi_new_data
difference_in_days
binary_df
```

# Another  sub step within the first part is to do a selection of some of the key partition in this data, identify and convert them to the long format.
# Our Partition of interest is the Response variables and the mean response variable , which is within the biger dataframe long_new,
# We will consider this variable as the Y which is been measured as a result of the  X but within the X variable we will partion again to examine 
# whether there are interventions, durations and patients because what we know is that modelling this factors together together can result in a model that we can fit to measure the change over time. 

# Here we are going to subset the Y variables which is the Response 
```{r}

# We set what we know as the Response variables and sebset them from the main data
response_columns<- c("DEJONGscore", "DEJONG_FUp1_Score", "DEJONG_FUp2_Score", 
                       "DEJONG_FUp3_Score", "DEJONG_FUp4_Score", "DEJONG_FUp5_Score",
                       "UCLA1Average", "UCLA_Fup1_Average", "UCLA_FUp2_Average", 
                       "UCLA_FUp3_Average", "UCLA_FUp4_Average", "UCLA_FUp5_Average",
                       "EQVASScore", "EQVAS_FUp1_score", "EQVAS_FUp2_score", 
                       "EQVAS_FUp3_score", "EQVAS_FUp4_score", "EQVAS_FUp5_score",
                       "SWEMWBSScore", "SWEMWBS_FUp1_score_num", "SWEMWBS_FUp2_score_num", 
                       "SWEMWBS_FUp3_score_num", "SWEMWBS_FUp4_score_num", "SWEMWBS_FUp5_score_num")

# Subset the data frame
Response <- longi_new_data %>% select(all_of(response_columns))

# View the subsetted data frame
print(Response)
```

# There are four coumns within the Y  therefore the mean response of each of the column or variable is identified from the dataset 
#  
```{r}
# We set what we know as the Response variables and sebset them from the main data
mean_response_columns<- c("DEJONGscore","UCLA1Average","EQVASScore","SWEMWBSScore")

# Subset the data frame
mean_response<- longi_new_data %>% select(all_of(mean_response_columns))

# View the subsetted data frame
print(mean_response)
```

```{r}
# Reshape data from wide to long format
Response_long_data <- Response %>%
  pivot_longer(
    cols = everything(), # Convert all columns
    names_to = "Measurement",
    values_to = "Score"
  ) %>%
  mutate(
    Year = case_when(
      str_detect(Measurement, "FUp1|Fup1") ~ "Year 1",
      str_detect(Measurement, "FUp2|Fup2") ~ "Year 2",
      str_detect(Measurement, "FUp3|Fup3") ~ "Year 3",
      str_detect(Measurement, "FUp4|Fup4") ~ "Year 4",
      str_detect(Measurement, "FUp5|Fup5") ~ "Year 5",
      TRUE ~ "Baseline"  # Default to Baseline if no FUp pattern matched
    )
  )

# View reshaped data
print(Response_long_data)
```
# Response = Patients + Duration + Interventions 
# There could be covariates or cofactors within that influence the interventions 
# longi_new_data,difference_in_days, binary_df, Response
# Patient + Intervention = longi_new_data - (difference_in_days + binary_df + Response)
# We rename Patient + Intervention as PI_data
# therefore Y = Response variable and X = PI_data + difference_in_days + binary_df , or X= longi_new_data-Response


```{r}
#Patient + Intervention = longi_new_data - (difference_in_days + binary_df + Response)

# Remove the selected columns from the original dataframe
# Corrected code
# Assuming difference_in_days, binary_df, and Response are already loaded as dataframes

# Extract column names from the three dataframes
columns_to_remove <- c(names(difference_in_days), names(binary_df), names(Response))

# Remove these columns from longi_new_data
PI_data <- longi_new_data %>%
  select(-all_of(columns_to_remove))

# Check the structure of the updated PI_data
str(PI_data)

# Check the new dataframe
head(PI_data)


```
# How to choose a model in Longitudinal data Machine learing, the first thing to do is to undertand the Response variable, 
# We have to find out if this data violates the model assumptions 
# we have to understand if the response variable which is mutivarate in this situation is correlated to one another,
# We have to verify if what type of data is in the response variable , we have to find out if it is a combination of categorical and numerical type of data in the response variable. 
# To choose a model we can use any of the following in different situations 
# lme4 is used for numerical contineous outcomes
# geepack for GEE or lme4 with a binomial family is used for binary classification outcome
# while mmm package is used for outcome that have contineous and categorical we can say mixed outcome but in situation where we have a complex hierarchical structure we make use of MCMCglmm model


```{r}
# we find out about the Response variable to understand why variables are present in the response variable, we use code and visualization to acheive this 
sapply(Response, class)
# We have all as interger except one of the response "SWEMWBSScore" which is numeric variable, therefore we can consider all Response variable as numeric variable but we will conver this distribution in a visualization.
```


```{r}
# Assuming your response dataframe is named `Response`
# Identify columns that are numeric or integer
numeric_columns <- sapply(Response, function(x) is.numeric(x) || is.integer(x))

# Filter the dataframe to keep only numeric and integer columns
numeric_df <- Response[, numeric_columns]

# Plot histograms for each numeric and integer column
for (col_name in names(numeric_df)) {
  p <- ggplot(numeric_df, aes_string(x = col_name)) +
    geom_histogram(binwidth = 5, fill = "blue", color = "black") +
    labs(title = paste("Histogram of", col_name), x = col_name, y = "Count") +
    theme_minimal()
  
  print(p)  # Correctly print the plot inside the loop
}

```
# Here we can see that DEJONG and UCLA are tending towards categorical in nature, maybe because of the way the data was preprocessed we are seeing a contineous data type been attached to them but in reality they are categorical, so what will do is to result to the data dictionary and find out what they are and then bring them back to their original data type which we believe is categorical and a factor. 

# The EQVAS and SWEMWBS are obviously contineous variable. 

```{r}

# List of columns to convert to categorical (factor)
columns_to_convert <- c("DEJONGscore", "DEJONG_FUp1_Score", "DEJONG_FUp2_Score", 
                        "DEJONG_FUp3_Score", "DEJONG_FUp4_Score", "DEJONG_FUp5_Score",
                        "UCLA1Average", "UCLA_Fup1_Average", "UCLA_FUp2_Average", 
                        "UCLA_FUp3_Average", "UCLA_FUp4_Average", "UCLA_FUp5_Average")

# Convert specified columns to factors (categorical variables)
Response_con <- Response %>%
  mutate_at(vars(one_of(columns_to_convert)), as.factor)

# Check the structure of the modified Response dataframe to ensure the conversion
str(Response_con)


# Convert specified columns to factors (categorical variables)
longi_new_data_con <- longi_new_data %>%
  mutate_at(vars(one_of(columns_to_convert)), as.factor)

# Check the structure of the modified Response dataframe to ensure the conversion
str(longi_new_data_con)
```



# In this type of response variable we are going to 
```{r}
# Load necessary library
library(ggplot2)

# Identify columns that are numeric or integer
numeric_columns <- sapply(Response_con, function(x) is.numeric(x) || is.integer(x))

# Identify columns that are categorical (factors or character)
categorical_columns <- sapply(Response_con, function(x) is.factor(x) || is.character(x))

# Filter the dataframe to keep only numeric columns
numeric_df <- Response_con[, numeric_columns]

# Filter the dataframe to keep only categorical columns
categorical_df <- Response_con[, categorical_columns]

# Plot histograms for each numeric and integer column
for (col_name in names(numeric_df)) {
  p <- ggplot(numeric_df, aes_string(x = col_name)) +
    geom_histogram(binwidth = 5, fill = "blue", color = "black") +
    labs(title = paste("Histogram of", col_name), x = col_name, y = "Count") +
    theme_minimal()
  
  print(p)  # Correctly print the plot inside the loop
}

# Plot bar charts for each categorical column
for (col_name in names(categorical_df)) {
  p <- ggplot(categorical_df, aes_string(x = col_name)) +
    geom_bar(fill = "lightgreen", color = "darkgreen") +
    labs(title = paste("Bar Chart of", col_name), x = col_name, y = "Count") +
    theme_minimal()
  
  print(p)  # Correctly print the plot inside the loop
}

```


```{r}
# Count unique values in each column
unique_countsR <- sapply(Response, function(x) length(unique(x)))

# Display the results
unique_countsR
```


```{r}
# Count unique values in each column
unique_countsR_c <- sapply(Response_con, function(x) length(unique(x)))

# Display the results
unique_countsR_c
```


```{r}
# We can conclude that this Response outcome is a Mix dataframe with muti categorical varables and Numerical variable 
# Therefore we will test our model asumptions in this direction. 
# The model of choice in this data will be 
```





```{r}


# Create a new dataframe that is a subset of Response_con
mean_response_df <- Response_con[, mean_response_columns]

# View the structure of the new dataframe to confirm
str(mean_response_df)
```


```{r}
# we proceed to remove the mean Response column to allow our data have the same lenght 

# Remove the specified columns from Response_con
Response_con_MR <- Response_con[, !names(Response_con) %in% mean_response_columns]

# View the updated dataframe
str(Response_con_MR)  # Check the structure to confirm the columns have been removed

```

```{r}
# We use the 

# Identify logical columns in the dataframe
logical_columns <- names(longi_new_data_con)[sapply(longi_new_data_con, is.logical)]

# Convert logical columns to factors
longi_new_data_con[logical_columns] <- lapply(longi_new_data_con[logical_columns], as.factor)

# View the structure of the updated dataframe to confirm the conversion
str(longi_new_data_con)
```

```{r}


# Identify numeric and integer columns (continuous variables)
numeric_columns <- names(longi_new_data_con)[sapply(longi_new_data_con, is.numeric)]

# Identify binary columns (factors with two levels)
binary_columns <- names(longi_new_data_con)[sapply(longi_new_data_con, function(x) is.factor(x) && length(levels(x)) == 2)]

# Identify categorical columns with more than two levels
categorical_columns <- names(longi_new_data_con)[sapply(longi_new_data_con, function(x) is.factor(x) && length(levels(x)) > 2)]

# Print the identified columns
cat("Numeric columns:", numeric_columns, "\n")
cat("Binary columns:", binary_columns, "\n")
cat("Categorical columns:", categorical_columns, "\n")

```


# There is a problem from here because i am trying to test for the assumption so i can appy the right model to the code, I might have to go back and use the Response_long but the issue is that it has all as numeric varable and if i try to convert the new columns with numeric and factors into a long format from Response_con , i get an error , so I try to split it as factors and numeric.
```{r}

# Define the Response_con variables
response_con_vars <- c("DEJONGscore", "DEJONG_FUp1_Score", "DEJONG_FUp2_Score", 
                       "DEJONG_FUp3_Score", "DEJONG_FUp4_Score", "DEJONG_FUp5_Score",
                       "UCLA1Average", "UCLA_Fup1_Average", "UCLA_FUp2_Average", 
                       "UCLA_FUp3_Average", "UCLA_FUp4_Average", "UCLA_FUp5_Average", 
                       "EQVASScore", "EQVAS_FUp1_score", "EQVAS_FUp2_score", 
                       "EQVAS_FUp3_score", "EQVAS_FUp4_score", "EQVAS_FUp5_score", 
                       "SWEMWBSScore", "SWEMWBS_FUp1_score_num", "SWEMWBS_FUp2_score_num", 
                       "SWEMWBS_FUp3_score_num", "SWEMWBS_FUp4_score_num", "SWEMWBS_FUp5_score_num")

# Separate numeric and factor (categorical) variables
numeric_responses <- response_con_vars[sapply(longi_new_data_con[response_con_vars], is.numeric)]
factor_responses <- response_con_vars[sapply(longi_new_data_con[response_con_vars], is.factor)]

# Check if there are any columns in each type
if (length(numeric_responses) > 0) {
  # Pivot numeric columns to long format
  numeric_long <- longi_new_data_con %>%
    select(all_of(numeric_responses)) %>%
    pivot_longer(
      cols = everything(),
      names_to = c(".value", "Year"),
      names_pattern = "(.*)_(FUp\\d*)?$"
    ) %>%
    mutate(
      Year = recode(Year, 
                    "FUp1" = "Year1", 
                    "FUp2" = "Year2", 
                    "FUp3" = "Year3", 
                    "FUp4" = "Year4", 
                    "FUp5" = "Year5"),
      Year = factor(Year, levels = c("Year1", "Year2", "Year3", "Year4", "Year5"))
    )
}

if (length(factor_responses) > 0) {
  # Pivot factor columns to long format
  factor_long <- longi_new_data_con %>%
    select(all_of(factor_responses)) %>%
    pivot_longer(
      cols = everything(),
      names_to = c(".value", "Year"),
      names_pattern = "(.*)_(FUp\\d*)?$"
    ) %>%
    mutate(
      Year = recode(Year, 
                    "FUp1" = "Year1", 
                    "FUp2" = "Year2", 
                    "FUp3" = "Year3", 
                    "FUp4" = "Year4", 
                    "FUp5" = "Year5"),
      Year = factor(Year, levels = c("Year1", "Year2", "Year3", "Year4", "Year5"))
    )
}

# Optional: Combine the long format data if both exist
if (exists("numeric_long") & exists("factor_long")) {
  combined_long <- bind_rows(numeric_long, factor_long)
} else if (exists("numeric_long")) {
  combined_long <- numeric_long
} else if (exists("factor_long")) {
  combined_long <- factor_long
}

# Check the resulting long format data
head(combined_long)

# Fit the model and test assumptions for numeric responses
if (exists("numeric_long")) {
  cat("\nTesting model assumptions for numeric response variables...\n")
  
  # Fit a linear mixed-effects model for one of the numeric responses, e.g., DEJONGscore
  numeric_model <- lmer(DEJONGscore ~ Year + (1 | ID), data = numeric_long)
  
  # Diagnostic plots to check assumptions
  par(mfrow = c(2, 2))
  plot(numeric_model)
  
  # Check for normality of residuals
  shapiro_test_result <- shapiro.test(residuals(numeric_model))
  cat("Shapiro-Wilk test for normality of residuals:\n")
  print(shapiro_test_result)
}

# Fit the model and test assumptions for factor (categorical) responses
if (exists("factor_long")) {
  cat("\nTesting model assumptions for factor response variables...\n")
  
  # Fit a generalized linear mixed-effects model for one of the factor responses
  factor_model <- glmer(as.factor(DEJONG_FUp1_Score) ~ Year + (1 | ID), 
                        data = factor_long, family = binomial)
  
  # Summary of the GLMM model
  cat("Summary of the GLMM model:\n")
  summary(factor_model)
}

```








```{r}

```




```{r}

```



```{r}

```



```{r}

```


```{r}

```


```{r}

```


```{r}

```




```{r}

```


```{r}

```



```{r}

```


```{r}

```



```{r}

```



```{r}

```